---
title: "Native figure and table numbering with knitr hooks"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Native figure and table numbering with knitr hooks}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

Quarto numbers figures and tables as static text ("Figure 1") and resolves `@fig-x` references
to static text too. `oq_numbering()` gives the figures and tables created by **R chunks** real
Word captions instead: `Figure <SEQ field>: caption`, wrapped in a bookmark named after the chunk
label. Word computes the numbers itself and renumbers when you insert or move items.

This is an opt-in alternative to the post-render option `officequarto.crossref.auto-number`
(`vignette("options")`), which converts Quarto's finished static captions afterwards. The two
differ in where they act:

| | `oq_numbering()` (this vignette) | `crossref.auto-number` |
|---|---|---|
| Acts | while knitting (knitr hooks) | after rendering (on the docx) |
| Reference syntax | <code>&#96;r oq_ref("fig-x")&#96;</code> (inline R) | `@fig-x` |
| Covers | figures/tables from R chunks | any Quarto-numbered caption |
| Customisable | your own hook function | fixed |

## Setup

Call it once in the setup chunk, after loading the package (hooks only exist inside the R session
that knits the document):

```r
library(officequarto)
oq_numbering()
```

Then write figures and tables as usual, with a `fig-`/`tbl-` label and a caption:

````{verbatim, lang = "markdown"}
```{r}
#| label: fig-sales
#| fig-cap: "Sales by region"
plot(sales)
```

```{r}
#| label: tbl-top
#| tbl-cap: "Top *five* customers"
knitr::kable(head(customers, 5))
```

As `r oq_ref("fig-sales")` and `r oq_ref("tbl-top")` show, ...
````

`oq_ref()` inserts a clickable Word `REF` field that displays the label and number, for example
"Figure 1" — what `@fig-sales` would render. For other output formats it returns `@fig-sales`, so
the same document still renders to HTML.

## Label text and separator

The label, separator and (for tables) caption style are chunk options, so they can be set once for
the whole project in `_quarto.yml`:

```yaml
knitr:
  opts_chunk:
    oq.fig.label: "Abbildung"
    oq.tbl.label: "Tabelle"
    oq.sep: ": "
    oq.tbl.style: "Table Caption"   # display name of the caption paragraph style
```

or overridden in a single chunk (`#| oq.fig.label: "Plate"`). Table captions may contain inline
markdown.

## A custom hook

`fig_hook` and `tbl_hook` replace the built-in post-processing. They are still chained onto
Quarto's own hooks, so they receive what Quarto produced and return the markdown to use instead:

- `fig_hook(res, options)` — `res` is the image markdown from Quarto's `plot` hook:
  `![caption](path){#fig-id width=...}`.
- `tbl_hook(res, options)` — `res` is the whole cell markdown from Quarto's `chunk` hook:
  `::: {#tbl-id .cell tbl-cap='...'} ...`.

`options` is the list of chunk options (`options$label`, `options$fig.cap`, your own `oq.*`
options, ...). The defaults are available as `oq_fig_hook()` and `oq_tbl_hook()`, so a custom hook
can extend them instead of starting from scratch:

```r
oq_numbering(
  fig_hook = function(res, options) {
    paste0(oq_fig_hook()(res, options), "\n\nSource: ", options$source, "\n\n")
  }
)
```

A custom hook that builds its own captions is responsible for the cross-reference id (drop `#fig-id`
from the image attributes, as the default does, if you do not want Quarto to number the figure as
well) and for unique bookmark names if you want `oq_ref()` to point at them.

## Limits

- Only R-chunk figures and tables are handled (label starting with `fig-`/`tbl-` plus
  `fig-cap`/`tbl-cap`). Markdown images and tables, sub-figures (`fig-subcap`) and `layout` chunks
  are left to Quarto.
- Quarto no longer knows the numbered items, so `@fig-x` / `@tbl-x` do **not** resolve for them —
  use `oq_ref()`. There is no list of figures or tables.
- Output other than docx is passed through unchanged.
- With `officequarto.crossref.auto-number: true` the hook-owned captions are skipped by the
  post-render conversion (no double conversion; checked in the test fixture).
- Table captions go in the paragraph style `Table Caption` (`oq.tbl.style`); your `reference-doc`
  should define it, otherwise Pandoc's fallback style is pruned and Word's default is used.
